---
title: "Homework 1.2 Report"
output:
  
  pdf_document: 
    df_print: kable
  word_document: default
  html_document: default
---
<center>
#### Monika Wysoczanska, 180817
#### Manuel Barbas, 180832
#### Diogo Oliveira, 180832\
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library("kableExtra")
library("moments")
library(pastecs)
library(ggplot2)
library("dplyr")
library(GGally)
```
##Wine Data Set


```{r, include=FALSE}

wines<-read.table("wines-PCA.txt", col.names = c("fixed acidity", "volatile acidity", "citic acid", "residual sugar", "chlorides", "free sulfur dioxide", "total sulfur dioxide", "density", "pH", "sulphates", "alcohol", "quality", "type"))
wines<-na.omit(wines)
summary(wines)
str(wines)

```
We have 98 observations of different wines, same number for both types: white and red. Each observation described by 12 quantitative variables and one categorical. 


```{r, include=FALSE}
wines$type<-as.factor(wines$type)
wines.pc<- select(wines, -quality)


#covariance matrix
cor(select(wines.pc, -type))
```
We observe very different variances, because of that for further analysis we use correlation matrix.

```{r, include=FALSE}
#correlation matrix)
cor(select(wines.pc, -type))
```

```{r, echo=FALSE}

ggpairs(wines.pc, lower = list(continuous="points",combo="facetdensity",mapping=aes(color=type)))

```
In the combined plot above we already see, that:
1. The data is not probably normally distributed among nearly all of the variables (except fixed acidity).
2. The variance of density is very small, then we cannot observe any significant correlation with other variables. Moreover, we are likely to deal with the outlier. 
We investigated the outlier a little bit deeper, and found that there is only one observation with a very high density value (10.02), which is:
```{r, include=FALSE}
boxplot(wines$density)
which(wines$density>1.5)

```
```{r}
wines[15,]

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
